Ordinal bhsdtr tutorial

### 2020-02-17 ordinal/Readme.Rmd
the third layer

This document contains a short tutorial on fitting hierarchical
ordinal polytomous regression models using the bhsdtr package, but
there is more; It also contains some explanations, illustrated with
annotated code, of several issues related to ordinal models in
general, and Signal Detection Theory and Item Response Theory in
particular. If you are interested in SDT, IRT, or ordinal models in
general, perhaps you will find something useful here.

@@

Ordinal polytomous variables are common in psychology, they are
usually analyzed by fitting a linear model, and all sorts of bad
things happen because of this, but only a few people seem to care. I
for one do not always care about this. We will not address this kind
of problems here, because we will analyze ordinal data using
(hierarchical) ordinal models.

@@

```{r echo=T,results="show"}
knitr::opts_chunk$set(echo=TRUE,results="show")
library(ggplot2)
library(bhsdtr)
library(Hmisc)
library(coda)
library(rstan)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
## This variable indicates if the fitted models can be loaded from disk
fresh_start = FALSE
sumfun = function(x)c(quantile(x, c(.025, .975)), mean(x))[c(1,3,2)]
```

@@

The number of parameters in this model matches the degrees of freedom
in the data, which means that the model is saturated, which means that
it always fits the data, but it does not mean that there is no
uncertainty associated with the estimates of its parameters.

@@

We cannot naively model the effects in the thresholds, because the
thresholds have to stay ordered, and assuming, for example, that each
threshold has a normal distribution due to the effect of participants
would violate the ordering assumption.

To supplement our model with a hierarchical linear regression
structure the vector of the thresholds has to be translated to the
unconstrained vector gamma. The 'log_distance' order-preserving link
function is one solution to this problem, but other order-preserving
link functions are also implemented in the bhsdtr package. The choice
of the link function may sometimes be important, but, since we will
not introduce numerical predictors for the thresholds, we can choose
any link function which seems most convenient.

The log_distance function is convenient for our purposes because it is
simple, and because the elements of the gamma vector under this link
function are easy to interpret and may be directly related to simple
research questions. In particular, one of the elements of the gamma
vector under the log_distance link function, i.e., the middle
threshold, will soon become important. This link function represents
the main threshold/criterion directly as a free parameter (here
gamma[4], because K = 8, and 8 / 2 = 4); Every other criterion is
represented as log of distance between the criterion and the one
criterion adjacent to it which is closer to the main criterion. This
way the order is preserved, and each threshold *i* is associated with
the corresponding unconstrained element *i* of the gamma
vector. Random and fixed effects in the thresholds are represented as
(hierarchical) linear effects in gamma. The two representations are
isomorphic, which means that we can always translate between them if
we need to.

@@

To fit a model we have to aggregate the data as much as possible
without distorting the hierarchical structure (this makes the process
of sampling much faster), create the fixed- and random-effects lists,
and create the data structures and the model code required by
stan. Note that we introduce a new parameter - eta - which represents
the average internal value. If we are going to model all the
thresholds (gamma parameter vector) then this parameter is redundant,
because a constant shift in the thresholds is equivalent to the
opposite effect in eta. The bhsdtr package will automatically fix the
first element (here the only element) of the eta fixed effects vector
at 0, which will make this model identifiable.

First, we will fit a model which assumes that the pattern of the
thresholds is the same for all the participants, but the mean of the
latent distribution may differ across participants. If this makes
little sense to you please take a look at the picture of the ordinal
model with ordered thresholds and imagine that the underlying
distribution has a different mean for different participants.

Note that the two assumptions, i.e., that the thresholds are constant
across participants and latent means or samples are participant-, not
item-specific seem to be often accepted in applications of Item
Response Theory. This is also reflected in the IRT terminology:
thresholds are often called "item-parameters" and latent means or
samples are often called "person-parameters".

@@

```{r echo=T,results="show"}
sdt_ml_plot(model, adata, sdata)
```

@@

```{r echo=T,results="show"}                
adata0 = aggregate_responses(d1, response = 'r', variables = c('id'))
## The only element of the eta fixed effects vector will be automatically fixed at 0
fixed0 = list(eta = ~ 1, gamma = ~ 1)
## Each participant has a unique patterh of the thresholds
random0 = list(list(group = ~ id, gamma = ~ 1))
model0 = make_stan_model(random0, model = 'ordinal', gamma_link = link)
sdata0 = make_stan_data(adata0, fixed0, random0, model = 'ordinal', gamma_link = link)
if(fresh_start){
    fit0 = stan(model_code = model0, data = sdata0,
                pars = c('gamma_fixed', 'gamma_random_1', 'gamma_sd_1'),
                init_r = .5,
                iter = 5000,
                chains = 4)
    saveRDS(fit0, '~/windows/temp/ordinal_fit_0.rds')
}else{
    fit0 = readRDS('~/windows/temp/ordinal_fit_0.rds')
}
```

@@

Now we will fit a model which is similar to an SDT model, but there
will be two important differences. We will not assume that the answer
"signal" cannot be less likely for the stimulus "signal" than for the
stimulus "noise", i.e., we will not assume that *d'* must be
non-negative, and we will *not* assume that the pattern of the
thresholds is the same for "noise" and "signal" stimuli. If you know
your Signal Detection Theory well, this may seem to you like a
perverse idea. If that is the case, please bear with me.

In this model each participant may have two different sets of
thresholds, one for "noise" and one for "signal" stimuli. Since the
thresholds are represented as random effects, we will account for the
fact that "random" thresholds may be correlated; They have to be
correlated, because if some threshold is higher (lower) than the
sample average, then the threshold above (below) it may also be higher
(lower).

@@

Now we can translate between the gamma and the criteria vectors (fixed
effects) for each stimulus:

@@

Note the difference between the main criterion (row 4). This
difference approximates *d'*. Even though, as we will later see, this
model does not correctly account for the effect of the stimulus
(because it violates the non-negativity assumption), we can still
recover this effect pretty well, because *d'* is equivalent to a
constant shift of all the thresholds in one predefined
direction. Shifting all the criteria to the right is equivalent to
shifting the underlying distribution of "latent values" (or evidence)
to the left, so we have to subtract the average position of the
criteria for the "signal" stimuli from the average position of the
criteria for the "noise" stimuli. Here is our first attempt:

@@

In an SDT model we assume that the stimulus class affects *only* the
*d'* parameter, and that both criteria (gamma) and *d'* may differ
between the participants (or items, etc.). In bhsdtr *d'* is modelled
as *log(d') = delta*, which forces it to be non-negative. This is an
important assumption of Signal Detection Theory - if this assumption
is violated, for example, because of the response reversal in some
participants or conditions, then an SDT model is not valid and a
non-trivial generalization of SDT should be used instead (see the
bhsdtr preprint for a more detailed explanation and some
references). Otherwise, this is the same general hierarchical ordinal
regression model.

```{r echo=T,results="show"}   
fixed2 = list(delta = ~ 1, gamma = ~ 1)
random2 = list(list(group = ~ id, delta = ~ 1, gamma = ~ 1))
```

Because of the importance of the stimulus class ("noise", or 1, and
"signal", or 2) in SDT models we have to indicate which variable
represents this information ( aggregate_responses(..., stimulus =
'stim', ...) ).

@@

The estimates match poorly. Or do they? What about the individual *d'*
estimates? To calculate the individual shifts in the thresholds /
criteria we have to add the random effects to the fixed effect:

@@

The point and interval estimates match quite well, with some notable
exceptions. According to the ordinal model some credible intervals
around individual *d'* point estimates include 0 and one point
estimate is even below 0, which contradicts an important assumption of
the SDT model. We will address this issue later.

@@

The SDT model seems to fit very well in a sense that the observed
distributions of responses are well within the 95% posterior
predictive intervals for each participant. However, this does not mean
much; an SDT model with multiple criteria is not saturated, so it does
not have to fit the data, but it does not leave too many degrees of
freedom in the data, so we can safely assume a priori that it will
usually fit the data well, regardless if it is true or not.

The discrepancy between the average *d'* estimates from both models is
not really a consequence of allowing for negative *d'* values in the
ordinal model. This discrepancy is a consequence of the assumed
distribution of individual shifts / *d'* values. In the ordinal model
we assumed that the shifts (*= d'*) are normally distributed, whereas
in the SDT model we assumed that *log(d')* are normally
distributed. We can see if the assumption that *log(d')* (or *d'*) are
normally distributed is approximately correct in our dataset. First,
let's take a look at participant-specific delta estimates in the SDT
model:

```{r echo=T,results="show"}
delta2 = apply(s2[['delta_fixed[1,1]']] + s2[, grep('delta_random_1', names(s2))], 2, mean)
ggplot(data.frame(delta = delta2), aes(sample = delta)) +
    stat_qq() +
    ylab('Sample quantile of log(d\') = delta random effects') +
    xlab('Theoretical normal quantile') +
    labs(title = 'SDT model')
```

The individual *delta = log(d')* estimates in the SDT model seem to be
approximately normally distributed. We cannot plot *log(d')* estimates
based on the ordinal model, because according to this model one
participant has a negative *d'* value and even without this
participant's data there are quite a few negative posterior *d'*
samples. However, we can see if the *d'* point estimates based on the
ordinal model are approximately normally distributed:

@@

They seem to be approximately normally distributed as well. The
average / fixed effect *d'* in each model is estimated as the mean of
different distributions: in the ordinal model participant-specific
shifts are assumed to be normally distributed and in the SDT model
*delta = log(d')* random effects are assumed to be normally
distributed. We can easily account for this by making use of the
relationship between the parameters of the normal distribution and the
associated log-normal distribution:

@@

Now we see that the two models give very similar point and interval
estimates of average *d'*.

@@

An SDT model is not only a statistical model, it also has *causal*
assumptions. Because the stimulus class was chosen at random on every
trial, every estimate of the statistical effect of the stimulus class
on any parameter of the model is an unbiased estimate of the total
causal effect of the stimulus class on this parameter. Moreover, in
such designs every statistical effect of the participant factor is an
estimate of the property or the causal effect of the
participants. This means that both the individual *d'* parameters and
the individual thresholds are participant-, not item-, parameters in
this model. So far so good.

However, in the SDT model we have also assumed that the stimulus class
can *only* affect the distance between the evidence distribution means
(= *d'*), whereas in the ordinal model we allowed for the possibility
of the stimulus class having an effect on the thresholds. It may seem
odd to you to allow for this, but there is nothing in the design of
the study that justifies the causal assumption of selective
influence. The estimates of the thresholds are based on the data that
were generated after the stimulus was presented, and so,
theoretically, the *pattern* of the thresholds could be affected by
the stimulus class as well.

It does not matter if an SDT model with selectivity assumption fits
the data well, because if the effect of the stimulus class is not
selective, the model is simply false. In particular, when the
selectivity assumption is false, an SDT model cannot deconfound
sensitivity / *d'* from bias / threshold / criteria effects.

The selectivity assumption is central to Signal Detection Theory. If
you know your Signal Detection Theory well, please take a moment and
try to recall one instant of this important assumption being
explicitly addressed in a paper or a book on Signal Detection Theory.

Before we go any further, just for fun, let's see what happens if we
introduce yet another assumption of selective influence, one that is
often accepted in Item Response Theory. We will fit a hierarchical SDT
model in which we will treat the thresholds as "item-parameters" and
we will treat the *d'* parameter as a "person-parameter".

```{r echo=T,results="show"}
fixed3 = list(delta = ~ 1, gamma = ~ 1)
random3 = list(list(group = ~ id, delta = ~ 1))
adata3 = aggregate_responses(d2, stimulus = 'stim', response = 'r', variables = c('id'))
sdata3 = make_stan_data(adata3, fixed3, random3, model = 'sdt', gamma_link = link)
model3 = make_stan_model(random3, model = 'sdt', gamma_link = link)
if(fresh_start){
    fit3 = stan(model_code = model3, data = sdata3,
                pars = c('delta_fixed', 'delta_sd_1', 'gamma_fixed', 'delta_random_1',
                         ## We will need this for the plot_sdt_fit function
                         'counts_new'),
                init_r = .5,
                iter = 5000,
                chains = 4)
    saveRDS(fit3, '~/windows/temp/ordinal_fit_3.rds')
}else{
    fit3 = readRDS('~/windows/temp/ordinal_fit_3.rds')
}
print(fit3, probs = c(.025, .975), pars = c('delta_fixed', 'gamma_fixed', 'delta_sd_1'))
```

@@

... we see that the simplified model not only gives biased
"person-parameter" estimates but the estimates seem more reliable
(here by about 20%). It does not make sense to look at the estimates
of the "item-parameters" since we already know that they have little
in common with the true thresholds.

@@

Note that we do not use the separate intercepts and slopes
parametrization here (i.e., ~ -1 + stim), because we are very much
interested in the *effect* of the stimulus class on the *pattern* of
the thresholds.

```{r echo=T,results="show"}   
fixed4 = list(eta = ~ 1, gamma = ~ stim)
random4 = list(list(group = ~ id, gamma = ~ stim))
adata4 = aggregate_responses(d2, response = 'r', variables = c('id', 'stim'))
sdata4 = make_stan_data(adata4, fixed4, random4, model = 'ordinal', gamma_link = link)
model4 = make_stan_model(random4, model = 'ordinal', gamma_link = link)
if(fresh_start){
    fit4 = stan(model_code = model4, data = sdata4,
                pars = c('gamma_fixed', 'gamma_random_1', 'gamma_sd_1'),
                init_r = .5,
                iter = 5000,
                chains = 4)
    saveRDS(fit4, '~/windows/temp/ordinal_fit_4.rds')
}else{
    fit4 = readRDS('~/windows/temp/ordinal_fit_4.rds')
}
print(fit4, probs = c(.025, .975), pars = c('gamma_fixed', 'gamma_sd_1'))
```

Here we see the estimated average shift / *d'* thought of as the mean
of normally distributed *d'* (not *delta = log(d')*) random effects
again:

@@

Ok, we did simplify things, but it still makes sense to take a look at
the other gamma effects, because these other effects correspond to the
variability in the *pattern* of the thresholds, and the estimates of
the thresholds in our ordinal model seem perfectly fine.

@@

Among the parameters that represent the stimulus-induced average
differences between the gamma parameters (i.e., gamma_fixed[.,2]) the
gamma_fixed[4,2] parameter does not seem to be the only one that is
non-zero. This indicates that the pattern of the thresholds - not only
their average position - is affected by the stimulus
class. Interestingly, the random effects' standard deviations tell a
different story:

@@

Note the clear pattern for the gamma_sd_1[.,2] parameters. Even though
the *average* pattern of the thresholds seems to be different
for the two stimulus classes (gamma_fixed[.,2] except for
gamma_fixed[4,2] ~ d'), this difference in pattern seems to be fairly
constant across the participants: all the gamma_sd_1[.,2] parameters
seem to be near-zero, except for the gamma_sd_1[4,2] parameter, which
corresponds to the variability in *d'* values. This is weird, since
the pattern of the thresholds for the "noise" stimuli clearly varies
with participants - none of the 95% HPD intervals for the
gamma_sd_1[.,1] parameters include zero: in terms of standard
deviations of posterior distributions of standard deviations (phew)
most are in fact quite far from zero.

Now we will modify the stan model code to account for the
non-negativity of *d'*. Let's index the rows of our dataset by
*i*. Because of the way the model was parametrized...

@@

when stim = "noise" = 0, gamma[4]_i = gamma_fixed[4,1] +
gamma_random_1[g_i][4,1], where g_i is the participant index in row
*i*. We can leave this part of the model intact, since this is the
main criterion position, which is an unconstrained parameter, but
also, as we can see below, the observed distribution of the main
criterion random effects is approximately normal:

@@

However, when stim = "signal" = 1, gamma[4]_i = gamma_fixed[4,1] +
gamma_random_1[g][4,1] + gamma_fixed[4,2] + gamma_random_1[g][4,2]. We
want this other part (the last sum which represents the difference in
gamma[4] between the "noise" and "signal" stimuli) to represent log of
average distance between the thresholds = log(shift) = log(d'), so we
have to subtract it from gamma[4]_i, exponentiate this sum, and add it
to gamma[4]_i.

@@

This is close. Not the same, but the model is different. What about
the bias = main criterion relative the the midpoint between the main
criteria?
```{r echo=T,results="show"}   
round(cbind(sumfun(s5[['gamma_fixed[4,1]']] - exp(s5[['gamma_fixed[4,2]']]) / 2),
      sumfun(s2[['gamma_fixed[4,1]']])), 2)
```

@@

We see again that gamma_sd_1[.,2] parameters, except for
gamma_sd_1[4,2] = log(d'), may be zero - the effect of the stimulus on
all the log-distances seems to be more or less the same for each
participant. What is going on here?

@@

...then after applying a non-linear transformation to the dependent
variable you may obtain a significant interacive effect:

@@

This means that whenever you see an interactive effect it may be an
artefact of the scale of the dependent variable. So, instead of
examining the effects in the gamma parameters. let's have a look at
the *average criteria / thresholds*. Notice that we have to add the
effect on gamma to gamma for stimulus = "noise" to obtain gamma for
stimulus = "signal".

@@

Now we can directly test if the average pattern of the criteria may be
different by comparing *centered* criteria vectors:

@@

When we center the criteria for the two classes of stimuli we no
longer see any evidence of the differences in their relative
positions. The only thing that seems to differ, and not by much, is
the main criterion, but this, I think, is confounded with *d'* in our
model. Selectivity seems to hold, kind of.

@@

>>>>

This document contains a short tutorial on fitting hierarchical
ordinal polytomous regression models using the bhsdtr package, but
there is more; It also contains some explanations, illustrated with
annotated code, of several issues related to ordinal models in
general, and Signal Detection Theory and Item Response Theory in
particular. If you are interested in any of these topics, perhaps you
will find something useful here.

@@

Ordinal polytomous variables are common in psychology, they are
usually analyzed by fitting a linear model, and all sorts of bad
things happen because of this, but most people do not seem to care. I
for one do not always care about this. We will not address this kind
of problems here, because we will analyze ordinal data using
(hierarchical) ordinal models.

@@

```{r echo=T,results="show"}
knitr::opts_chunk$set(echo=TRUE,results="show")
library(ggplot2)
library(bhsdtr)
library(Hmisc)
library(coda)
library(rstan)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
## This variable indicates if the fitted models can be loaded from
## disk. I use it for caching
fresh_start = FALSE
sumfun = function(x)c(quantile(x, c(.025, .975)), mean(x))[c(1,3,2)]
```

@@

The number of parameters in this model matches the degrees of freedom
in the data (if the data are aggregated), which means that the model
is saturated, which means that it always fits the data, but it does
not mean that there is no uncertainty associated with the estimates of
its parameters.

@@

We cannot naively model the effects in the thresholds, because the
thresholds have to stay ordered, and assuming, for example, that each
threshold has a normal distribution due to the effect of participants
would violate the ordering assumption. To supplement our model with a
hierarchical linear regression structure the vector of the thresholds
has to be translated to the unconstrained vector gamma. The
'log_distance' order-preserving link function is one solution to this
problem, but other order-preserving link functions are also
implemented in the bhsdtr package. The choice of the link function may
sometimes be important, but, since we will not introduce quantitative
predictors for the thresholds, we can choose any link function which
seems most convenient.

The log_distance function is convenient for our purposes because it is
simple, and because the elements of the gamma vector under this link
function are easy to interpret and may be directly related to simple
research questions. This link function represents the main threshold /
criterion directly as a free parameter (here gamma[4], because K = 8,
and 8 / 2 = 4); Every other criterion is represented as log of
distance between the criterion and the one criterion adjacent to it
which is closer to the main criterion. This way the order is
preserved, and each threshold *i* is associated with the corresponding
unconstrained element *i* of the gamma vector. Random and fixed
effects in the thresholds are represented as (hierarchical) linear
effects in gamma. The two representations are isomorphic, which means
that we can always translate between the gamma and the ordered vectors
if we need to.

@@

To fit a model we have to aggregate the data as much as possible
without distorting the hierarchical structure (this makes the process
of sampling much faster), create the fixed- and random-effects lists,
and create the data structures and the model code required by
stan. Note that we introduce a new parameter - eta - which represents
the average internal value. If we are going to model the effects in
all the thresholds (gamma parameter vector) then this parameter is
redundant, because a constant shift in the thresholds is equivalent to
the opposite effect in eta. The bhsdtr package will automatically fix
the first element (here the only element) of the eta fixed effects
vector at 0, which will make this model identifiable.

First, we will fit the model which assumes that the pattern of the
thresholds is the same for all the participants, but the mean of the
latent distribution may differ across participants. If this makes
little sense to you please take a look at the picture of the ordinal
model with ordered thresholds and imagine that the underlying
distribution has a different mean for different participants.

Note that the two assumptions, i.e., that the thresholds are constant
across participants and that the latent means or samples are
participant-, not item-specific seem to be often accepted in
applications of Item Response Theory. This is also reflected in the
IRT terminology: thresholds are often called "item-parameters" and
latent means or samples are often called
"person-parameters".

@@

```{r echo=T,results="show"}
## This function is not defined in the bhsdtr package
sdt_ml_plot(model, adata, sdata)
```

@@

```{r echo=T,results="show"}                
adata0 = aggregate_responses(d1, response = 'r', variables = c('id'))
## The only element of the eta fixed effects vector will be automatically fixed at 0
fixed0 = list(eta = ~ 1, gamma = ~ 1)
## Each participant has a unique patterh of the thresholds / gamma
random0 = list(list(group = ~ id, gamma = ~ 1))
model0 = make_stan_model(random0, model = 'ordinal', gamma_link = link)
sdata0 = make_stan_data(adata0, fixed0, random0, model = 'ordinal', gamma_link = link)
if(fresh_start){
    fit0 = stan(model_code = model0, data = sdata0,
                pars = c('gamma_fixed', 'gamma_random_1', 'gamma_sd_1'),
                init_r = .5,
                iter = 5000,
                chains = 4)
    saveRDS(fit0, '~/windows/temp/ordinal_fit_0.rds')
}else{
    fit0 = readRDS('~/windows/temp/ordinal_fit_0.rds')
}
```

@@

Now we will fit a model which is similar to an SDT model, but there
will be two important differences. We will not assume that the answer
"signal" cannot be less likely for the stimulus "signal" than for the
stimulus "noise", i.e., we will not assume that *d'* must be
non-negative, and we will *not* assume that the pattern of the
thresholds is the same for the "noise" and the "signal" stimuli. If
you know your Signal Detection Theory well, this may seem to you like
a perverse idea. If that is the case, please bear with me.

In this model each participant may have two different sets of
thresholds, one for the "noise" and one for the "signal" stimuli. We
will account for the fact that participant specific threshold effects
may be correlated; They are usually correlated, because if some
threshold is higher (lower) than the sample average, then the
threshold above (below) it may also be higher (lower).

@@

Now we can again translate between the gamma and the criteria vectors
(fixed effects) for each stimulus:

@@

Note the difference between the main criterion (row 4). This
difference approximates *d'*. Even though, as we will later see, this
model does not correctly account for the effect of the stimulus
(because it violates the non-negativity assumption), we can still
recover this effect pretty well, because *d'* is equivalent to a
constant shift of all the thresholds in one predefined
direction. Shifting all the criteria to the right is equivalent to
shifting the underlying distribution of latent values (or evidence) to
the left, so we have to subtract the average position of the criteria
for the "signal" stimuli from the average position of the criteria for
the "noise" stimuli. Here is our first attempt:

@@

In an SDT model we usually assume that the stimulus class affects
*only* the *d'* parameter, and that both criteria (gamma) and *d'* may
differ between the participants (or items, etc.). In bhsdtr *d'* is
represented as *log(d') = delta*, which forces it to be
non-negative. This is an important assumption of Signal Detection
Theory - if this assumption is violated, for example, because of the
response reversal in some participants or conditions, then an SDT
model is not valid and a non-trivial generalization of SDT should be
used instead (see the bhsdtr preprint for a more detailed explanation
and some references).

```{r echo=T,results="show"}   
## We want one average d' / delta and one vector of average
## thresholds / criteria / gammma
fixed2 = list(delta = ~ 1, gamma = ~ 1)
## Each participant has a unique d' / delta and a unique set of
## thresholds / criteria / gamma
random2 = list(list(group = ~ id, delta = ~ 1, gamma = ~ 1))
```

Because of the importance of the stimulus class in SDT models
("noise", or 1, and "signal", or 2) we have to indicate which variable
represents this information ( aggregate_responses(..., stimulus =
'stim', ...) ) when calling the aggregate_responses function.

@@

The estimates match poorly. Or do they? What about the individual *d'*
estimates? To calculate the individual shifts in the thresholds /
criteria in the more general ordinal model we have to add the random
effects of participants to the fixed effect:

@@

The point and interval estimates match quite well (*r* = .93 for the
point estimates), with some notable exceptions. According to the
ordinal model some credible intervals around individual *d'* point
estimates include 0 and one point estimate is even below 0, which
contradicts an important assumption of the SDT model. We will address
this issue later.

@@

The SDT model seems to fit very well in a sense that the observed
distributions of responses are well within the 95% posterior
predictive intervals for each participant. However, this does not mean
much; An SDT model with multiple criteria is not saturated, so it does
not have to fit the data, but it does not leave too many degrees of
freedom in the data either. We can safely assume a priori that it will
usually fit the data well, regardless if it is true or not.

The discrepancy between the average *d'* estimates from both models is
not really a consequence of allowing for negative *d'* values in the
ordinal model. This discrepancy is a consequence of the assumed
distribution of individual shifts / *d'* values. In the ordinal model
we assumed that the shifts (*= d'*) are normally distributed, whereas
in the SDT model we assumed that *log(d')* are normally
distributed. We can see if the assumption that *log(d')* (or *d'*) are
normally distributed is approximately correct in our dataset. First,
let's take a look at the participant-specific delta estimates in the
SDT model:

```{r echo=T,results="show"}
## Participant's delta = average delta + participant's effect
delta2 = apply(s2[['delta_fixed[1,1]']] + s2[, grep('delta_random_1', names(s2))], 2, mean)
ggplot(data.frame(delta = delta2), aes(sample = delta)) +
    stat_qq() +
    ylab('Sample quantile of log(d\') = delta random effects') +
    xlab('Theoretical normal quantile') +
    labs(title = 'SDT model')
```

The individual *delta = log(d')* estimates in the SDT model seem to be
approximately normally distributed. We cannot plot *log(d')* estimates
based on the ordinal model, because according to this model one
participant has a negative *d'* value and even without this
participant's data there are quite a few negative posterior *d'*
samples. However, we can see if the *d'* point estimates based on the
general ordinal model are approximately normally distributed:

@@

They seem to be approximately normally distributed as well. The
average *d'* in each model is estimated as a mean of different
distributions: in the ordinal model participant-specific shifts are
assumed to be normally distributed, but in the SDT model *delta =
log(d')* random effects are assumed to be normally distributed. We can
easily account for this by making use of the relationship between the
parameters of the normal distribution and the associated log-normal
distribution:

@@

Now we see that the two models give very similar point and interval
estimates of average *d'*. How cool is this?

@@

# Causal assumptions of Signal Detection Theory

An SDT model is not only a statistical model, it also has *causal*
assumptions. To be clear, an SDT model *as a statistical model* does
not have any causal assumptions. However, causal assumptions are
present when the parameters of this model are interpreted as
properties of a psychological process, or when it is used to
*deconfound* (a causal notion) sensitivity (another extra-statistical
notion) from bias (this term has a technical meaning in statistics,
but it means something *psychological* in an SDT model - *a tendency
to prefer one class of stimuli over the other*).

Because the gabor dataset comes from a study in which the stimulus
class was chosen at random on every trial, every estimate of the
statistical effect of the stimulus class on any parameter of the model
is an unbiased estimate of the total causal effect of the stimulus
class on this parameter. Moreover, in such designs every statistical
effect of the participant factor is an estimate of the property or the
causal effect of the participants. This means that both the individual
*d'* parameters and the individual thresholds are participant-, not
item-, parameters in this model. So far so good.

However, in the SDT model we also assumed that the stimulus class can
*only* affect the distance between the evidence distribution means (=
*d'*), whereas in the ordinal model we allowed for the possibility of
the stimulus class having an effect on the pattern of the
thresholds. It may seem odd to you to allow for this, but there is
nothing in the design of the study that justifies the causal
assumption of selective influence. The estimates of the thresholds are
based on the data that were generated after the stimulus was
presented, and so, theoretically, the pattern of the thresholds could
be affected by the stimulus class as well.

It does not matter if an SDT model with selectivity assumption fits
the data well, because if the effect of the stimulus class is not
selective, the model is simply false. In particular, when the
selectivity assumption is false, an SDT model cannot deconfound
sensitivity / *d'* from bias / threshold / criteria effects. Once we
drop the selectivity assumption there is no longer any *d'* / internal
value parameter, only the thresholds, like in our general ordinal
model. The selectivity assumption is central to Signal Detection
Theory. If you know your Signal Detection Theory well, please take a
moment and try to recall one instant of this important assumption
being explicitly addressed in a paper or a book on Signal Detection
Theory. If you manage to find one such example, please send me an
email.

Before we go any further, just for fun, let's see what happens if we
introduce yet another assumption of selective influence. We will fit a
hierarchical SDT model in which we will treat the thresholds as
"item-parameters" and we will treat the *d'* parameter as a
"person-parameter".

```{r echo=T,results="show"}
fixed3 = list(delta = ~ 1, gamma = ~ 1)
## Each participant has a unique d', but not thresholds
random3 = list(list(group = ~ id, delta = ~ 1))
adata3 = aggregate_responses(d2, stimulus = 'stim', response = 'r', variables = c('id'))
sdata3 = make_stan_data(adata3, fixed3, random3, model = 'sdt', gamma_link = link)
model3 = make_stan_model(random3, model = 'sdt', gamma_link = link)
if(fresh_start){
    fit3 = stan(model_code = model3, data = sdata3,
                pars = c('delta_fixed', 'delta_sd_1', 'gamma_fixed', 'delta_random_1',
                         ## We will need this for the plot_sdt_fit function
                         'counts_new'),
                init_r = .5,
                iter = 5000,
                chains = 4)
    saveRDS(fit3, '~/windows/temp/ordinal_fit_3.rds')
}else{
    fit3 = readRDS('~/windows/temp/ordinal_fit_3.rds')
}
print(fit3, probs = c(.025, .975), pars = c('delta_fixed', 'gamma_fixed', 'delta_sd_1'))
```

@@

... we see that the simplified model not only gives biased
"person-parameter" estimates but the estimates seem more reliable
(they are shorter by about 20% on average). It does not make sense to
look at the estimates of the "item-parameters" since we already know
that they have little in common with the true thresholds.

@@

Note that we do not use the separate intercepts and slopes
parametrization here (we use ~ stim instead of ~ -1 + stim), because
we are very much interested in the difference in the pattern of the
thresholds caused by the stimulus class.

```{r echo=T,results="show"}   
## Separate thresholds for "signa" and "noise"
fixed4 = list(eta = ~ 1, gamma = ~ stim)
## Each participants has two unique sets of thresholds, one for "signa" and one for "noise".
random4 = list(list(group = ~ id, gamma = ~ stim))
adata4 = aggregate_responses(d2, response = 'r', variables = c('id', 'stim'))
sdata4 = make_stan_data(adata4, fixed4, random4, model = 'ordinal', gamma_link = link)
model4 = make_stan_model(random4, model = 'ordinal', gamma_link = link)
if(fresh_start){
    fit4 = stan(model_code = model4, data = sdata4,
                pars = c('gamma_fixed', 'gamma_random_1', 'gamma_sd_1'),
                init_r = .5,
                iter = 5000,
                chains = 4)
    saveRDS(fit4, '~/windows/temp/ordinal_fit_4.rds')
}else{
    fit4 = readRDS('~/windows/temp/ordinal_fit_4.rds')
}
print(fit4, probs = c(.025, .975), pars = c('gamma_fixed', 'gamma_sd_1'))
```

Here we see the estimated average shift / *d'* thought of as the mean
of the normally distributed *d'* (not *delta = log(d')*) random
effects again:

@@

Ok, we did simplify things a bit, but it still makes sense to take a
look at the other gamma effects, because these other effects
correspond to the variability in the *pattern* of the thresholds, and
the estimates of the thresholds in our ordinal model seem perfectly
fine.

@@

Among the parameters that represent the stimulus-induced average
differences between the gamma parameters (i.e., gamma_fixed[.,2]) the
gamma_fixed[4,2] parameter does not seem to be the only one that is
non-zero. This indicates that the pattern of the thresholds - not only
their average position (= gamma_fixed[4,.]) - may be affected by the
stimulus class. Interestingly, the random effects' standard deviations
tell a different story:

@@

Note the clear pattern for the gamma_sd_1[.,2] parameters. The
*average* pattern of the thresholds seems to be different for the two
stimulus classes (gamma_fixed[.,2] except for gamma_fixed[4,2] ~ d'),
and participants' thresholds for the "noise" stimuli seem to differ
(gamma_sd_1[.,1]) - none of the 95% HPD intervals for the
gamma_sd_1[.,1] parameters include zero: in terms of standard
deviations of posterior distributions of standard deviations (phew)
most are in fact quite far from zero. However, the difference in the
pattern of the thresholds for the "signal" and the "noise" stimuli
seems to be fairly constant across the participants: all the
gamma_sd_1[.,2] parameters seem to be near-zero, except for the
gamma_sd_1[4,2] parameter, which corresponds to the variability in the
*d'* parameter. This is weird.

We will modify the stan model code to account for the non-negativity
of *d'*. Let's index the rows of our dataset by *i*. Because of the
way the model was parametrized...

@@

... when stim = "noise" = 0, gamma[4]_i = gamma_fixed[4,1] +
gamma_random_1[g_i][4,1], where g_i is the participant index in row
*i*. We can leave this part of the model intact, since this is the
main criterion position, which is an unconstrained parameter, but
also, as we can see below, the observed distribution of the main
criterion random effects is approximately normal:

@@

When stim = "signal" = 1, gamma[4]_i = gamma_fixed[4,1] +
gamma_random_1[g][4,1] + gamma_fixed[4,2] + gamma_random_1[g][4,2]. We
want this other part (the last sum which represents the difference in
gamma[4] between the "noise" and the "signal" stimuli) to represent
log of average distance between the thresholds = log(shift) = log(d'),
so we have to subtract it from gamma[4]_i, exponentiate this sum, and
add it to gamma[4]_i.

@@

This is close. Not the same, but the model is different. What about
the bias = main criterion relative the the midpoint between the main
criteria?

```{r echo=T,results="show"}   
round(cbind(sumfun(s5[['gamma_fixed[4,1]']] - exp(s5[['gamma_fixed[4,2]']]) / 2),
      sumfun(s2[['gamma_fixed[4,1]']])), 2)
```

@@

We see again that the gamma_sd_1[.,2] parameters, except for
gamma_sd_1[4,2] = log(d'), may be zero - the effect of the stimulus on
all the log-distances seems to be more or less the same for each
participant. What is going on here?

@@

...then after applying a non-linear transformation to the dependent
variable you may observe a significant interacive effect:

@@

This means that whenever you see an interactive effect it may be an
artefact of the scale of the dependent variable. So, instead of
examining the effects in the gamma parameters. let's have a look at
the average *criteria / thresholds*. Notice that we have to add the
effect on gamma to gamma for stimulus = "noise" to obtain gamma for
stimulus = "signal".

@@

Now we can directly test if the average pattern of the criteria may be
different by comparing the *centered* criteria vectors:

@@

When we center the criteria for the two classes of stimuli we no
longer see any evidence of the differences in their relative
positions. The only thing that seems to differ, and not by much, is
the main criterion, but this, I think, is confounded with *d'* in our
model (this is only my guess). Selectivity seems to hold, kind of. To
really justify the selectivity assumption we have to argue for the
null, which, in a bayesian setting, usually means using the bayes
factor. Even though I love bayesian inference, I do not like the bayes
factor, and I am not alone in this.

@@

# The missing part

At the risk of sounding ridiculous, I will now make use of the fact
that I can say whatever I want in an informal tutorial such as this
one. Also, I just had a beer.

It may seem that once (maybe in some distant future) we solve the
statistical problems and the causal problems of testing an SDT model
we can say that we are done. We have the great formal tools of
statistical and causal inference at our disposal, and they are
indispensable, but there is one last layer of abstraction that we did
not explicitly identify.
